---
title: "Homework #2"
author: Nicholas Matthews; nickmatthews713
date: "2/3/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This homework is about syntax and basic plotting functions in R, and is **due by the end of Feb 7**. 

You can use the RMarkDown file with these homework instructions as a template for your work. 
Make sure to replace "Your Name" and "your_github_username" in the YAML with your name and Github username.

**Submission instructions**: Create a folder named `ds202_hw2`, and name the RMarkDown file including your solutions `hw2.Rmd` under this folder. For submission, create a GitHub repository named `ds202_hw2` under your account, and push both `hw2.Rmd` and the knitted `hw2.html` to GitHub before the deadline. I will assume you use the same GitHub username as for your HW1 submission. The latest version of your homework appearing on GitHub before the deadline is graded. *It is crucial to follow the exact folder structure and file names*, so that your homework can be reproduced and graded by an automated script. Some questions will require the following answer format.

## Example question and answer
0\. What is the smallest positive odd integer? Storage it in a variable named `a0` and print it.

**Example answer**:
In the RMarkDown file, type in
<pre><code>```{r}
a0 <- 1
a0
```</code></pre>
The knitted html file will show:
```{r}
a0 <- 1
a0
```

## R commands
1. Today's temperature has a high of 24 and a low of 15 degree fahrenheit.  Find the corresponding degress in celsius and storage it in a (single) vector named `cel`. Print the variable.

```{r}

fer <- c(24,15)
cel <- (fer-32)*(5/9)
cel

```

2. Create a vector of length 200. The odd entries should be 3 and even entries should be 1. Storage the created in a variable named `vec200` and print.

```{r}

vec200 <- rep(c(1,3), length.out=200)
vec200

```
3. How many perfect squares are there between 1 and 2001? Storage the answer in a  variable named `numPerfect` and print. You may want to look at `?round`.

```{r}

numPerfect <- floor(sqrt(2001)) - ceiling(sqrt(1)) + 1
numPerfect

```

## Cars and mileage
For this part we will use the `mpg` data set, you can access the data by first installing (using `install.packages('ggplot2')`) and then loading the package `ggplot2` (using `library(ggplot2)`) into your R session. The `mpg` data includes information about fuel economy of popular car models
for the years 1999 and 2008 as collected by the US Environmental Protection Agency. An answer should include a short paragraph of text for the explanation to a (human) reader, together with appropriate R code, outputs, and graphs.

Most exercises are from the ggplot2 book section 2.2.1. You might find it helpful for working on the homework to read through this section. *However, all questions can be solved by using just the basic R graphics functions.*

1. What are the 3 most fuel-efficient models in terms of highway mileage? Storage the data frame (with 3 rows) containing all information of these three cars in a variable named `top3` and print.

```{r}
library(ggplot2)

car_efficient <- mpg[order(mpg$hwy, decreasing = TRUE),]
top3 <- car_efficient[1:3,]
top3

```
##### description
I used 'library(ggplot)' to include the data frame mpg in my Rmarkdown file. Then, I ordered the dataset by it highway miles per gallon. To do this, I indexed the rows of the data set with a vector returned by the function 'order()' and kept all columns included. This is stored inside of the car_efficient variable. I used indexing once again to store the top 3 rows from car_efficient into variable top3  

2. How many different compact models are included in this dataset? Storage the answer in  a variable named `numCompact` and print.

```{r}

compact <- mpg[mpg$class=='compact',]
numCompact <- table(compact$model)
numCompact

```
##### description
First, I filtered out the compact cars in the dataset by making a new dataset that only includes compact cars using indexing. Then, I used the 'table()' function to count the number of times each model appears within my new dataset.

3. Create a scatterplot of `hwy` and `cty` and describe the relationship. Why are there so few points visible? Explain your findings.

```{r}

ggplot(data = mpg) + geom_point(mapping=aes(x=hwy, y=cty))

```

##### description  

The scatterplot displays a positive correlation between highway miles per gallon and cities miles per gallon for each car. In other words, If the car has good highway miles per gallon, then the car will most likely have good cities miles per gallon. There are so little points because there is a lot of overlapping of points. For example, rows 6 and 8 of the original dataset both have a value of 16 for cities miles per gallon and a value of 26 for highway miles per gallon.

4. Are cars in 2008 more fuel efficient than cars in 1999? Use graphs *and* summary statistics to answer this question. Explain your findings.

```{r}

mpg2008 <- mpg[mpg$year==2008, c('hwy','cty')]
mpg1999 <- mpg[mpg$year==1999, c('hwy','cty')]
mpg2008$mpg.avg <- (mpg2008$hwy + mpg2008$cty)/2
mpg1999$mpg.avg <- (mpg1999$hwy + mpg1999$cty)/2

summary(mpg2008)
summary(mpg1999)

mpg$mpg.avg = (mpg$hwy + mpg$cty)/2

ggplot(data=mpg, aes(x=year, y=mpg.avg, group=year)) + geom_boxplot(fill="slateblue", alpha=0.2) + ylab("average miles per gallon")

```

##### description  

I started by making 2 data sets that include only the years 2008 and 1999  respectively. Then, to simplify and define the "fuel efficiency" of each car, I made an extra column for each set that takes the average of the cars highway miles per gallon and city miles per gallon. Then, by looking at the average of this column for both years, 1999 cars appear to be slightly more efficient.